---
title: "Predictions"
author: "Robert F Gruener"
date: "6/26/2021"
output: html_document
---


Expression uses an illumina probe. Looking at the [original publication](http://europepmc.org/article/MED/22522925#S8), the probeset is Illumina HT-12 v3


Install the illuminaHumanv3.db package to enable conversion

```{r}
#install or update these packages if needed:

#install.packages("BiocManager")
#BiocManager::install("illuminaHumanv3.db") #Bioconductor version '3.12' requires R version '4.0'; use `BiocManager::install(version = '3.13')` with R version 4.1; see https://bioconductor.org/install



```


Load libraries and read in Metabric Expression Data

```{r}

library(tidyverse)
library(illuminaHumanv3.db)

Metabric_expression <- read_csv("Metabric Data/Complete_normalized_expression_data_METABRIC.csv") #load in metabric data

Metabric_expression[1:5,1:5] #preview


```


```{r}
symbol_lookupup <- data.frame(Gene=unlist(mget(x = Metabric_expression$ILMN_Probe, envir = illuminaHumanv3SYMBOL))) #create a map for ID to symbols

table(is.na(symbol_lookupup$Gene)) #returns 20,138 NA values...

ensem_look <- data.frame(Gene=unlist(mget(x = Metabric_expression$ILMN_Probe, envir = illuminaHumanv3ENSEMBL)))

table(is.na(ensem_look$Gene)) #still returns 20,657 NA values... may need to look into, but it seems like these values don't map to genes surprisingly?

#ignoring any ensembl data for now


#the symbol map and the metabric expression data are the same length and same order, so add in the Gene column

Metabric_expression$Gene <- symbol_lookupup$Gene
Metabric_expression <- select(Metabric_expression, ILMN_Probe, Gene, everything())

Metabric_expression[1:5, 1:5] #preview
```




```{r}
#Filter the NA rows out of both

filtered_metabric <- drop_na(Metabric_expression, Gene)

#setting row names at this point doesnt't work
#For some reason, a lot of the illumina probes map on to the same gene
#row.names(filtered_metabric) <- symbol_lookup$Gene[!is.na(symbol_lookup$Gene)]

genes <- filtered_metabric$Gene
dups_genes <- genes[duplicated(genes)] #10217 genes are in the dataset twice
length(unique(genes)) #only 19,221 unique genes


#need to figure out it we should average the expression of these genes, chooose one, omit them completely...

#filtered_metabric[genes %in% dups_genes, ]

correlation_dupgenes <- numeric()
rmse_dupgenes <- numeric()
average_expression_x <- numeric()
average_expression_y <- numeric()

for(i in 1:length(dups_genes)){
  gene <- dups_genes[i]

x <- filtered_metabric[filtered_metabric$Gene == gene,][1, 3:ncol(filtered_metabric)] %>% 
  as.numeric()
y <- filtered_metabric[filtered_metabric$Gene == gene,][2, 3:ncol(filtered_metabric)] %>% 
  as.numeric()

correlation_dupgenes[i] <- cor(x, y)
rmse_dupgenes[i] <- ModelMetrics::rmse(x, y)

average_expression_x[i] <- mean(x)
average_expression_y[i] <- mean(y)
  
}
summary(correlation_dupgenes)

hist(correlation_dupgenes)
hist(rmse_dupgenes)

hist(average_expression_x)
hist(average_expression_y)

hist(average_expression_x - average_expression_y)


gene <- dups_genes[3]

x <- filtered_metabric[filtered_metabric$Gene == gene,][1, 3:ncol(filtered_metabric)] %>% 
  as.numeric()
y <- filtered_metabric[filtered_metabric$Gene == gene,][2, 3:ncol(filtered_metabric)] %>% 
  as.numeric()
z <- filtered_metabric[filtered_metabric$Gene == gene,][3, 3:ncol(filtered_metabric)] %>% 
  as.numeric()

plot(x, y)
plot(x, z)
plot(z, y)

cor(x, y)
cor(x, z)
cor(z, y)

```


Based on all the above, it is unclear why there are these duplicates and also which set of the duplicates to use if any. 

The correlation data indicates that most don't correlate well. This could have been becaues they don't correlate well or that this is an indication of low expression  (i.e. if a gene has low expression, the value might be driven by noise and therefore wouldn't correlate well). However, looking at the average expression, it seems that these genes are all expressed for the most part. 

Looking at the example plots at the end, it seems like these genes are are just similarly expressed in all the samples (pretty low variance, e.g. from 5-6 or 6-7).

Overall, we probably could just average and it would be fine. However, I'm going to pre-emptively remove the low correlation values because those in theory shouldn't be useful for the model anyway and we wouldn't know which value to trust. For those that correlate well, we will keep them. 

```{r}
table(correlation_dupgenes > 0.6) #keep about 2455 genes

to_keep <- dups_genes[correlation_dupgenes > 0.6]

temp_dup_metabric <- filtered_metabric %>% 
  filter(Gene %in% to_keep) %>% 
  group_by(Gene) %>% 
  summarise_if(is.numeric, mean)

metabric_expression_final <- filtered_metabric %>% 
  filter(!Gene %in% dups_genes) %>% #remove all duplicates
  select(-ILMN_Probe) %>%  #get rid of the extra column 
  bind_rows(., temp_dup_metabric) #add back the genes that we averaged above. 

```


Should be ready for some predictions now (hopefully)

```{r}
Response_data <- readRDS("Prepped_Data/CTRPv2_Response_Matrix.rds")
training_expression_data <- readRDS("Prepped_Data/CTRP_Expression_Matrix.rds")

convert.to.matrix <- function(df){
  temp <- as.matrix(df[,-1])
  row.names(temp) <- pull(df, 1)
  return(temp)
  
}
test_expression_metabric <- convert.to.matrix(metabric_expression_final)

set.seed(12345)
#install.packages("oncoPredict")


library(oncoPredict)

#to make this a little faster, I'm not running it with all drugs, just a few
ptype_info <- Response_data[, which(colnames(Response_data) %in% c("leptomycin B", "MK-1775", "tamoxifen"))]


oncoPredict::calcPhenotype(trainingExprData = training_expression_data, 
                           trainingPtype = ptype_info, 
                           testExprData = test_expression_metabric, 
                           batchCorrect = "rank", 
                           minNumSamples = 20, 
                           printOutput = T, 
                           removeLowVaringGenesFrom = "homogenizeData")


```


A folder should be created in the working directory titled calcPhenotype_Output which has a .csv file titled DrugPredictions.csv. We'll use this in the next script: Analyze_predictions.Rmd




